---
title: "cgMLST attribution experiments"
author: "Jonathan Marshall, Massey University"
date: "22 November 2019"
output: 
  ioslides_presentation: 
    widescreen: yes
    css: css/ioslides.css
    logo: logos/idrec_square.png
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(grid)
library(knitr)
theme_set(theme_bw(base_size=15))
opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align='center', bg='transparent', dev.args=list(bg='transparent'))

ans <- read_csv("~/cgmlst/confusion.csv")

best <- ans %>% filter(str_detect(Genes, "Best"), str_detect(Data, "CV")) %>%
  extract(Genes, into='Genes', regex="([0-9]+)", convert=TRUE)
mlst <- ans %>% filter(str_detect(Genes, "MLST") | str_detect(Genes, "Sheppard"),
                       str_detect(Data, "CV")) %>%
  mutate(Genes = case_when(Genes == "MLST" ~ 7,
                           Genes == "rMLST" ~ 52,
                           Genes == "Sheppard" ~ 13))
random <- ans %>% filter(str_detect(Genes, "Random_1[0-9]"), str_detect(Data, "CV"))

attr <- read_csv("~/cgmlst/attribution.csv")
best_attr <- attr %>% filter(str_detect(Genes, "Best"), str_detect(Data, "Full")) %>%
  extract(Genes, into='Genes', regex="([0-9]+)", convert=TRUE) %>%
  spread(quantile, value)
mlst_attr <- attr %>% filter(str_detect(Genes, "MLST") | str_detect(Genes, "Sheppard"),
                       str_detect(Data, "Full")) %>%
  mutate(Genes = case_when(Genes == "MLST" ~ 7,
                           Genes == "rMLST" ~ 52,
                           Genes == "Sheppard" ~ 13)) %>%
  spread(quantile, value)

random_attr <- attr %>% filter(str_detect(Genes, "Random"), str_detect(Data, "Full")) %>%
  extract(Genes, into="Genes", regex="([0-9]+)", convert=TRUE) %>%
  mutate(Genes = sprintf("R%02i", Genes)) %>%
  bind_rows(attr %>% filter(str_detect(Genes, "^MLST"), str_detect(Data, "Full"))) %>%
  spread(quantile, value)

fig_width = 10
```

## Choosing the best genes

Use the F-statistic:
$$
\begin{aligned}
F &= \frac{\mathsf{Between~source~variation}}{\mathsf{Within~source~variation}}\\
 &= \frac{\mathsf{Total~source~variation} - \mathsf{Within~source~variation}}{\mathsf{Within~source~variation}}
\end{aligned}
$$
We can estimate Total and Within source variation just as the sum of squared pairwise distances.

Similar to how MANOVA or PERMANOVA works.

## Choosing the best genes

- Choose the gene that has maximal F.

- Choose the next gene that, combined with the first, maximises F.

- Keep adding the next best gene that maximises F combined with the genes already chosen.

- Locally optimal, not necessarily globally optimal.

## cgMLST attribution

```{r}
ggplot(best_attr, aes(x=Genes, y=`0.5`, fill=Source)) +
  geom_ribbon(aes(ymin=`0.025`, ymax=`0.975`), alpha=0.3) +
  geom_ribbon(aes(ymin=`0.25`, ymax=`0.75`), alpha=0.3) +
  geom_line(aes(col=Source), lwd=2) +
  scale_y_continuous("Attribution of human cases", labels = scales::percent, expand=c(0,0), limits=c(0,1)) +
  scale_x_continuous(limits = c(0,100), expand=c(0,0)) +
  guides(col = 'none', fill='none')
```

## What about MLST, Sheppard, rMLST

- MLST: Standard 7 genes

- Sheppard: 13 genes selected via GWAS to separate sources in the UK.

    - Thépault et al (2017) Genome-wide identification of host-segregating epidemiological markers for source attribution in *Campylobacter jejuni*.

- rMLST: Ribosomal MLST.

## MLST, Sheppard, rMLST

```{r}
ggplot(best_attr, aes(x=Genes, y=`0.5`, fill=Source)) +
  geom_ribbon(aes(ymin=`0.025`, ymax=`0.975`), alpha=0.3) +
  geom_ribbon(aes(ymin=`0.25`, ymax=`0.75`), alpha=0.3) +
  geom_line(aes(col=Source), lwd=2) +
  geom_linerange(data=mlst_attr, aes(ymin=`0.025`, ymax=`0.975`), col='black') +
  geom_linerange(data=mlst_attr, aes(ymin=`0.25`, ymax=`0.75`), lwd=2) +
  geom_point(data=mlst_attr, size=2, shape=21, col='black') +
  scale_y_continuous("Attribution of human cases", labels = scales::percent, expand=c(0,0), limits=c(0,1)) +
  scale_x_continuous(limits = c(0,100), expand=c(0,0)) +
  guides(col = 'none', fill='none')
```

## Cross validation

- Use only the source isolates

- Divide into 10 groups, preserving source structure in each group.

- Use 9 to train the island model, and retain the last group as "Human" cases.

- Compare attributed source to actual source.

- Repeat across the 10 groups (cross-validation).

## Cross validation results

```{r, fig.width=10}
p <- ggplot(best, aes(x=Genes, y=Percent, col=Source)) + geom_line(aes(group=interaction(Source, Data)), alpha=0.5) +
  geom_line(data=best %>% group_by(Genes, TrueSource, Source) %>% summarise(Percent = mean(Percent)) %>% ungroup(), lwd=1.5) +
  facet_wrap(~TrueSource) + scale_y_continuous("Percent attribution", labels = scales::percent, expand=c(0,0), limits=c(0,1)) +
  scale_x_continuous(limits = c(0,100), expand=c(0,0)) +
  guides(col = 'none', fill='none')

g <- ggplot_gtable(ggplot_build(p))
facet_top <- which(grepl('strip-t', g$layout$name))
fills <- alpha(scales::hue_pal(h = c(0, 360) + 15, c = 100, l = 65)(3), 0.5)
k <- 1
for (i in facet_top) {
  j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
  k <- k+1
}
grid.draw(g)
```

## With MLST, Sheppard genes, rMLST

```{r}
p <- ggplot(best, aes(x=Genes, y=Percent, col=Source)) + geom_line(aes(group=interaction(Source, Data)), alpha=0.5) +
  geom_line(data=best %>% group_by(Genes, TrueSource, Source) %>% summarise(Percent = mean(Percent)) %>% ungroup(), lwd=1.5) +
  geom_point(data=mlst, aes(group=interaction(Source, Data)), alpha=0.5, size=2) +
  geom_point(data=mlst %>% group_by(Genes, TrueSource, Source) %>% summarise(Percent = mean(Percent)) %>% ungroup(), aes(fill=Source), size=3, shape=21, col='black') +
  facet_wrap(~TrueSource) + scale_y_continuous("Percent attribution", labels = scales::percent, expand=c(0,0), limits=c(0,1)) +
  scale_x_continuous(limits = c(0,100), expand=c(0,0)) +
  guides(col = 'none', fill='none')

g <- ggplot_gtable(ggplot_build(p))
facet_top <- which(grepl('strip-t', g$layout$name))
fills <- alpha(scales::hue_pal(h = c(0, 360) + 15, c = 100, l = 65)(3), 0.5)
k <- 1
for (i in facet_top) {
  j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
  k <- k+1
}
grid.draw(g)
```

## What about 7 genes picked at random?

```{r}
p <- ggplot(random, aes(x=Genes, y=Percent, col=Source)) + geom_boxplot() +
  #  geom_line(data=ans %>% group_by(Genes, true_source, Source) %>% summarise(Percent = mean(Percent)) %>% ungroup(), lwd=1.5) +
  facet_wrap(~TrueSource) + scale_y_continuous("Percent attribution", labels = scales::percent, expand=c(0,0), limits=c(0,1)) +
  #  scale_x_continuous(limits = c(0,100), expand=c(0,0)) +
  guides(col = 'none') +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank())

g <- ggplot_gtable(ggplot_build(p))
facet_top <- which(grepl('strip-t', g$layout$name))
fills <- alpha(scales::hue_pal(h = c(0, 360) + 15, c = 100, l = 65)(3), 0.5)
k <- 1
for (i in facet_top) {
  j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
  g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
  k <- k+1
}
grid.draw(g)
```

## Attribution from 7 random genes

```{r}
ggplot(random_attr, aes(x=Genes, y=`0.5`, fill=Source)) +
  geom_linerange(aes(ymin=`0.025`, ymax=`0.975`), col='black') +
  geom_linerange(aes(ymin=`0.25`, ymax=`0.75`, col=Source), lwd=2) +
  geom_point(size=2, shape=21, col='black') +
  scale_y_continuous("Attribution of human cases", labels = scales::percent, expand=c(0,0), limits=c(0,1)) +
 # scale_x_discrete(limits = c(0,100), expand=c(0,0)) +
  guides(col = 'none', fill='none') +
  theme(axis.title.x = element_blank())
```

## Conclusion

- Any 7 genes seem to do just as well as MLST.

- Increasing the number of genes appears to break the attribution model.

- The F-statistic doesn't appear to correspond well with how well attribution happens.

- Need a better measure? (And to fix the model!)
